---
title: "CRC Image Feature Analysis"
format:
  html:
    theme: yeti
    highlight-style: tango
    toc: true
    toc-depth: 3
    toc-floating: true
---

## Image Analysis: TCGAData

#### Setting up environment

```{r}
#| output: False
renv::init()
renv::snapshot()
renv::status()

```

#### Setting up libraries

```{r}
#| output: False
library(BiocManager)
library(MOFA2)
library(remotes)
library(SingleCellExperiment)
library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(MultiAssayExperiment)
library(SummarizedExperiment)
library(curatedTCGAData)
library(dplyr)
library(spatstat)
library(moments)
library(pracma)
library(HistoImageR)
library(devtools)
library(ggplot2)
library(TCGAutils)
library(rmarkdown)
library(knitr)
library(psych)

```

#### prov-gigaPath Colorectal Cancer WSI 

```{r}
getwd()
dir <- "../TCGA_COAD_READ_provgigapath"
fnames <- list.files(dir, pattern = "\\.csv$", full.names = TRUE)
length(fnames)  # how many slides

```

#### build embedding matrix

```{r}
#| label: build-embedding-matrix
#| cache: true
#| warning: false
#| message: false

# Load helper utils fns
devtools::load_all()
#------------------------------------------
# Simplify filenames to patient IDs (clean)
#------------------------------------------
participant_ids <- gsub(".csv$", "", basename(fnames)) %>% 
  strsplit(., "-") %>% 
  sapply(., function(x) paste(x[1:3], collapse = "-"))

# import WSI embedding layers 
structure_layer0  <- getEmbeddingLayer(fnames, participant_ids, layer = 0)
structure_layer10 <- getEmbeddingLayer(fnames, participant_ids, layer = 10)
structure_layer11 <- getEmbeddingLayer(fnames, participant_ids, layer = 11)
structure_layer12 <- getEmbeddingLayer(fnames, participant_ids, layer = 12)
structure_last <- getEmbeddingLayer(fnames, participant_ids, layer = "last_layer_embed" )


```



#### Matching embedding and assay data COAD + READ


```{r}
#------------------------------------------
# 1. Load COAD + READ RNASeq assays
#------------------------------------------
# Load COAD
coad_mae <- curatedTCGAData(
  diseaseCode = "COAD",
  assays = "RNASeq2GeneNorm",
  version = "2.1.1",
  dry.run = FALSE
)

# Load READ
read_mae <- curatedTCGAData(
  diseaseCode = "READ",
  assays = "RNASeq2GeneNorm",
  version = "2.1.1",
  dry.run = FALSE
)

rna_coad <- coad_mae[["COAD_RNASeq2GeneNorm-20160128"]]
rna_read <- read_mae[["READ_RNASeq2GeneNorm-20160128"]]


```
# Prepare summarise experiment objects

```{r}
#------------------------------------------
# 2. Ensure consistent genes and merge RNASeq
#------------------------------------------
common_genes <- intersect(rownames(rna_coad), rownames(rna_read))
rna_coad_bulk <- rna_coad[common_genes, , drop = FALSE]
rna_read_bulk <- rna_read[common_genes, , drop = FALSE]
rna_combined <- cbind(rna_coad_bulk, rna_read_bulk)

# Standardize patient IDs (first 3 TCGA fields)
rna_patients <- sapply(strsplit(colnames(rna_combined), "-"), function(x) paste(x[1:3], collapse = "-"))
colnames(rna_combined) <- make.unique(rna_patients)

#------------------------------------------
# 3. Prepare embedding layers (keep rownames as patient IDs)
#------------------------------------------
prepare_layer <- function(mat) {
  if (is.null(rownames(mat))) {
    rownames(mat) <- paste0("patient_", seq_len(nrow(mat)))
  }
  mat_unique <- mat[!duplicated(rownames(mat)), , drop = FALSE]
  return(mat_unique)
}

structure_layer0  <- prepare_layer(structure_layer0)
structure_layer10 <- prepare_layer(structure_layer10)
structure_layer11 <- prepare_layer(structure_layer11)
structure_layer12 <- prepare_layer(structure_layer12)
structure_last    <- prepare_layer(structure_last)

#------------------------------------------
# 4. Prepare clinical metadata
#------------------------------------------
coad_clinical <- as.data.frame(colData(coad_mae))
read_clinical <- as.data.frame(colData(read_mae))

surv_vars <- c("vital_status.x", "days_to_death.x", 
               "patient.days_to_last_followup", "patient.days_to_last_known_alive")

# Subset columns with >95% completeness or survival variables
coad_sub <- coad_clinical[, colSums(!is.na(coad_clinical))/nrow(coad_clinical) * 100 > 95 | colnames(coad_clinical) %in% surv_vars, drop = FALSE]
read_sub <- read_clinical[, colSums(!is.na(read_clinical))/nrow(read_clinical) * 100 > 95 | colnames(read_clinical) %in% surv_vars, drop = FALSE]

# Remove technical columns
coad_sub <- coad_sub %>% select(-matches("aliquot"), -matches("portion"))
read_sub <- read_sub %>% select(-matches("aliquot"), -matches("portion"))

# Harmonize columns across COAD + READ
all_cols <- union(colnames(coad_sub), colnames(read_sub))
for (col in setdiff(all_cols, colnames(coad_sub))) coad_sub[[col]] <- NA
for (col in setdiff(all_cols, colnames(read_sub))) read_sub[[col]] <- NA

clinical_sub <- rbind(coad_sub, read_sub)

# Factorize categorical variables (2–10 levels)
numLv <- sapply(clinical_sub, function(x) length(unique(x)))
slideAttrNames <- names(numLv[numLv >= 2 & numLv <= 10])
for (attr in slideAttrNames) clinical_sub[[attr]] <- as.factor(clinical_sub[[attr]])

#------------------------------------------
# 5. Match patients with embeddings
#------------------------------------------
patients_with_embed <- Reduce(union, lapply(list(structure_layer0, structure_layer10,
                                                 structure_layer11, structure_layer12,
                                                 structure_last), rownames))

# Subset clinical metadata to only patients with embeddings
metadata_df <- clinical_sub[rownames(clinical_sub) %in% patients_with_embed, , drop = FALSE]

#------------------------------------------
# 6. Match RNASeq to patients with embeddings
#------------------------------------------
# Subset to matching samples (columns)
rna_matched <- rna_combined[, colnames(rna_combined) %in% rownames(metadata_df)]

# Extract the actual numeric expression matrix from the assay
rna_matched_mat <- assay(rna_matched)

# Check dimensions
dim(rna_matched_mat)
# should be genes × samples, e.g., 17995 × 606

# Remove genes with any non-finite values (NA, Inf, -Inf)
rna_keep <- apply(rna_matched_mat, 1, function(x) all(is.finite(x)))
rna_matched_clean <- rna_matched_mat[rna_keep, ]

#------------------------------------------
# 7. Create SummarizedExperiment objects for each embedding layer
#------------------------------------------

# Ensure metadata rownames match patient IDs
rownames(metadata_df) <- metadata_df$patientID

make_embed_se <- function(mat, metadata_df) {
  # Keep only patients present in both the matrix and metadata
  patients <- intersect(rownames(mat), rownames(metadata_df))
  
  # Subset matrix and metadata
  mat_subset <- mat[patients, , drop = FALSE]
  metadata_subset <- metadata_df[patients, , drop = FALSE]
  
  # Create SE: columns = patients, rows = features/dims
  se <- SummarizedExperiment(
    assays = list(embedding = t(mat_subset)),  # t(): features x patients
    colData = metadata_subset
  )
  
  # Give rownames to features/dims
  rownames(se) <- paste0("dim_", seq_len(nrow(se)))
  return(se)
}

embedding_se_layer0  <- make_embed_se(structure_layer0, metadata_df)
embedding_se_layer10 <- make_embed_se(structure_layer10, metadata_df)
embedding_se_layer11 <- make_embed_se(structure_layer11, metadata_df)
embedding_se_layer12 <- make_embed_se(structure_layer12, metadata_df)
embedding_se_last    <- make_embed_se(structure_last, metadata_df)


#------------------------------------------
# 8. Create RNA SummarizedExperiment
#------------------------------------------
# Ensure colData rownames match assay column names
coldata_rna <- DataFrame(patientID = colnames(rna_matched_clean))
rownames(coldata_rna) <- colnames(rna_matched_clean)

rna_se <- SummarizedExperiment(
  assays = list(counts = assay(rna_matched_clean)),
  colData = coldata_rna
)


#------------------------------------------
# 9. Build MultiAssayExperiment
#------------------------------------------
# Make sure colData rownames match RNA columns
metadata_mae <- metadata_df[colnames(rna_se), , drop = FALSE]

crc_mae <- MultiAssayExperiment(
  experiments = list(
    RNASeq = rna_se,
    #mbedding_layer0  = embedding_se_layer0,
    #Embedding_layer10 = embedding_se_layer10,
    #Embedding_layer11 = embedding_se_layer11,
    #Embedding_layer12 = embedding_se_layer12,
    Embedding_last    = embedding_se_last
  ),
  colData = metadata_mae
)

# Check rownames of each experiment
exps <- experiments(crc_mae)
lapply(exps, function(x) head(rownames(x)))
```






#### Running MOFA2 analysis

```{r}
# Create MOFA object
mofa_object <- create_mofa_from_MultiAssayExperiment(crc_mae)
# Define model options
model_opts <- get_default_model_options(mofa_object)
model_opts$num_factors <- 15

# Define training options
train_opts <- get_default_training_options(mofa_object)
train_opts$convergence_mode <- "slow"
train_opts$seed <- 42

# Prepare and run MOFA
mofa_object <- prepare_mofa(
  object = mofa_object,
  model_options = model_opts,
  training_options = train_opts
)



```

```{r}
#| label: train-mofa-model

# Run/train the model 

mofa_object.trained <- run_mofa(
  mofa_object,
  outfile = "mofa_model_rnaseq_lastembed.hdf5",
  use_basilisk = TRUE
)

# Load trained MOFA model
mofa_object.trained <- load_model("mofa_model_rnaseq_lastembed.hdf5")

#------------------------------------------
# 1. Get MOFA sample names
#------------------------------------------
mofa_samples <- unlist(samples_names(mofa_object.trained))  # vector of sample IDs

#------------------------------------------
# 2. Subset metadata to match MOFA samples
#------------------------------------------
metadata_matched <- metadata_df[mofa_samples, , drop = FALSE]

# Ensure rownames match MOFA sample names exactly
rownames(metadata_matched) <- mofa_samples

#------------------------------------------
# 3. Add sample column if needed
#------------------------------------------
metadata_matched$sample <- rownames(metadata_matched)

#------------------------------------------
# 4. Assign metadata to MOFA object
#------------------------------------------
samples_metadata(mofa_object.trained) <- metadata_matched

# Save the updated object (RDS is easiest)
saveRDS(mofa_object.trained, file = "mofa_model_lastembed_with_metadata.rds")


# visualize the model views and samples
plot_data_overview(mofa_object.trained)

# Plot factor correlations to check for interdependence issues with the model
plot_factor_cor(mofa_object.trained)


# Plot variance explained
plot_variance_explained(mofa_object.trained)

plot_variance_explained(mofa_object.trained, plot_total = T)
print(mofa_object.trained@cache$variance_explained)


```



## misc

```{r}
# Extract embedding matrices
mat_layer0  <- assay(embedding_se_layer0, "embedding")
mat_layer10 <- assay(embedding_se_layer10, "embedding")
mat_layer11 <- assay(embedding_se_layer11, "embedding")
mat_layer12 <- assay(embedding_se_layer12, "embedding")
mat_last    <- assay(embedding_se_last, "embedding")

# Calculate correlation between same embedding dimensions across layers
layer_corr <- data.frame(
  dim = rownames(mat_layer0),
  cor_0_10  = diag(cor(t(mat_layer0), t(mat_layer10))),
  cor_10_11 = diag(cor(t(mat_layer10), t(mat_layer11))),
  cor_11_12 = diag(cor(t(mat_layer11), t(mat_layer12))),
  cor_12_last = diag(cor(t(mat_layer12), t(mat_last)))
)

head(layer_corr)
summary(layer_corr)
```


#### Load in TCGA CRC assay experiment data

```{r}
#| output: False
#| message: false

# List available Assays
curatedTCGAData("COAD", "*",version = "2.1.1", dry.run = TRUE) 
curatedTCGAData("READ", "*",version = "2.1.1", dry.run = TRUE) 


# Load COAD data with multiple assays
coad_mae_full <- curatedTCGAData(
    diseaseCode = "COAD",
    assays = c("COAD_RNASeq2GeneNorm-20160128", "mRNAArray", "miRNASeqGene"),
    version = "2.1.1",
    dry.run = FALSE
)

# Load READ data (rectal)
read_mae_full <- curatedTCGAData(
    diseaseCode = "READ",
    assays = c("READ_RNASeq2GeneNorm-20160128", "mRNAArray", "miRNASeqGene"),
    version = "2.1.1",
    dry.run = FALSE
)

```

#### Read Clinical Metadata data and subset

```{r}
#| output: False
# ---- Extract sample-level metadata for READ ----- 
read_clinical <- colData(read_mae_full)

# Check completeness of metadata attributes
completeness <- colSums(!is.na(read_clinical)) / nrow(read_clinical) * 100
summary(completeness)

# Subset to attributes with >95% completeness and remove technical fields
read_clinical_sub <- read_clinical[, which(completeness > 95)] %>%
  as.data.frame() %>%
  select(-matches("aliquot")) %>%
  select(-matches("portion"))


# ------ Extract sample-level metadata for COAD ------
coad_clinical <- colData(coad_mae_full)

# Check completeness of metadata attributes
completeness <- colSums(!is.na(coad_clinical)) / nrow(coad_clinical) * 100

coad_clinical_sub <- coad_clinical[, which(completeness > 95)] %>%
  as.data.frame() %>%
  select(-matches("aliquot")) %>%
  select(-matches("portion"))


```

#### Matching embedding with clinical metadata

```{r}
#| output: False
# Merge COAD + READ clinical metadata
all_cols <- union(colnames(coad_clinical_sub), colnames(read_clinical_sub))

coad_clinical_full <- coad_clinical_sub
for (col in setdiff(all_cols, colnames(coad_clinical_sub))) coad_clinical_full[[col]] <- NA

read_clinical_full <- read_clinical_sub
for (col in setdiff(all_cols, colnames(read_clinical_sub))) read_clinical_full[[col]] <- NA

clinical_combined <- rbind(coad_clinical_full, read_clinical_full)

# Keep only patients with embeddings
# Convert embedding matrix rownames to patient IDs (first 3 components)
patient_ids_from_embedding <- sapply(strsplit(rownames(tensor_matrix_full), "-"), 
                                    function(x) paste(x[1:3], collapse = "-"))
rownames(tensor_matrix_full) <- patient_ids_from_embedding

# Convert clinical data rownames to patient IDs (first 3 components)
clinical_patient_ids <- sapply(strsplit(rownames(clinical_combined), "-"), 
                              function(x) paste(x[1:3], collapse = "-"))
rownames(clinical_combined) <- clinical_patient_ids

# Remove duplicate patients from clinical data (keep first occurrence)
clinical_combined <- clinical_combined[!duplicated(rownames(clinical_combined)), ]

# Keep only patients with embeddings
patients_with_embed <- rownames(tensor_matrix_full)
clinical_combined <- clinical_combined[rownames(clinical_combined) %in% patients_with_embed, ]

# Factorize categorical variables with 2-10 unique values
numLv <- apply(clinical_combined, 2, function(x) length(unique(na.omit(x))))
slideAttrNames <- names(numLv[numLv >= 2 & numLv <= 10])

for(attr in slideAttrNames){
  clinical_combined[[attr]] <- as.factor(clinical_combined[[attr]])
}

# Create final metadata dataframe
metadata_embed_df <- clinical_combined[slideAttrNames]

# Remove any rows with all NA values
metadata_embed_df <- metadata_embed_df[rowSums(is.na(metadata_embed_df)) < ncol(metadata_embed_df), ]

# Subset embedding matrix to match metadata patients
embedding_embed_mat <- tensor_matrix_full[rownames(tensor_matrix_full) %in% rownames(metadata_embed_df), ]
embedding_embed_mat <- embedding_embed_mat[!duplicated(rownames(embedding_embed_mat)), ]

# Ensure exact same patients in both datasets
common_patients <- intersect(rownames(embedding_embed_mat), rownames(metadata_embed_df))
embedding_embed_mat <- embedding_embed_mat[common_patients, ]
metadata_embed_df <- metadata_embed_df[common_patients, ]

# Run HistoImageR embedding separation analysis
res_embed <- HistoImageR:::analyze_embedding_separation(embedding_embed_mat, metadata_embed_df)

# View results
res_embed$plots
```


