---
title: "embed_analysis"
format: html
editor: visual
---

## Libraries

```{r, message=FALSE}
library(rmarkdown)
library(knitr)
library(psych)
#BiocManager::install("waldronlab/imageFeatureTCGA")
packageVersion("imageFeatureTCGA")
library(imageFeatureTCGA)
library(curatedTCGAData)
library(SummarizedExperiment)
library(dplyr)
library(TCGAutils)
library(ggplot2)
library(pheatmap)
library(GSVA)
library(msigdbr)
library(tidyr)
library(patchwork)
library(survival)
library(ggrepel)
```


# Load slide data catalog (Kidney Renal Cancer)

```{r load embeddings}
cat =getCatalog("provgigapath") 
# Kidney
catalog_kidney <- getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-KIRC")


# lung
catalog_lung <- getCatalog("provgigapath") |> 
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-LUAD")


# Colorectal
catalog_coad <- getCatalog("provgigapath") |> 
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-COAD")


# Ovarian
catalog_ovar <- getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-OV")

# Rectal
catalog_rect <- getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-READ")



```


# Read Cancer assays with curatedTCGA

```{r load_rnaseq, message=FALSE}

# Load Kidney assays
kidney_mae <- curatedTCGAData(
    diseaseCode = "KIRC",
    assays =  c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
    version = "2.1.1",
    dry.run = FALSE
)

lung_mae <- curatedTCGAData(
  diseaseCode = "LUAD",
  assays = c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
  version = "2.1.1",
  dry.run = FALSE
)

COAD_mae <- curatedTCGAData(
  diseaseCode = "COAD",
  assays = c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
  version = "2.1.1",
  dry.run = FALSE
)


OV_mae <- curatedTCGAData(
  diseaseCode = "OV",
  assays =  c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
  version = "2.1.1",
  dry.run = FALSE
)


READ_mae <- curatedTCGAData(
  diseaseCode = "READ",
  assays = c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
  version = "2.1.1",
  dry.run = FALSE
)

# For OV
upsetSamples(ovarian_mae_emb, sets = c("slide_assay", "OV_RNASeq2GeneNorm-20160128", 
                                       "OV_RPPAArray-20160128"))
upsetSamples(kidney_mae_emb, sets = c("slide_assay", "KIRC_RNASeq2GeneNorm-20160128",
                                    "KIRC_RPPAArray-20160128", "KIRC_mRNAArray-20160128"))
# For READ
upsetSamples(read_mae_emb, sets = c("slide_assay", "READ_RNASeq2GeneNorm-20160128",
                                    "READ_RPPAArray-20160128", "READ_mRNAArray-20160128"))

```


# Align embeddings

### Helper Function

* Make sure there's no redundant samples
* Averages all samples belonging to one patient


```{r}
get_aligned_embedding_metadata <- function(mae_emb) {
  # mae_emb is output from linkTCGA (MultiAssayExperiment with slide_assay)
  # Align patients across assays and return embedding matrix (patients x features) and metadata
  
  # Add row names to slide_assay if not already
  if (is.null(rownames(mae_emb[["slide_assay"]]))) {
    rownames(mae_emb[["slide_assay"]]) <- paste0("slide_feat_", seq_len(nrow(mae_emb[["slide_assay"]])))
  }
  
  # Get patients with slide data
  slide_sm <- sampleMap(mae_emb)[sampleMap(mae_emb)$assay == "slide_assay", ]
  patients <- unique(slide_sm$primary)
  
  # Build patient-level matrices for each assay
  aligned <- list()
  for (nm in names(experiments(mae_emb))) {
    se <- experiments(mae_emb)[[nm]]
    mat <- assay(se)
    sm <- sampleMap(mae_emb)[sampleMap(mae_emb)$assay == nm, ]
    col_patients <- sm$primary[match(colnames(mat), sm$colname)]
    patient_mat <- matrix(NA, nrow = nrow(mat), ncol = length(patients),
                          dimnames = list(rownames(mat), patients))
    for (pat in patients) {
      cols <- which(col_patients == pat)
      if (length(cols) == 1) patient_mat[, pat] <- mat[, cols]
      else if (length(cols) > 1) patient_mat[, pat] <- rowMeans(mat[, cols, drop = FALSE], na.rm = TRUE)
    }
    if (!is.numeric(patient_mat)) {
      patient_mat <- matrix(as.numeric(patient_mat), nrow = nrow(patient_mat),
                            dimnames = dimnames(patient_mat))
    }
    keep <- rowSums(!is.na(patient_mat)) > 0
    aligned[[nm]] <- patient_mat[keep, , drop = FALSE]
  }
  
  # Embedding matrix (features x patients) -> transpose to patients x features
  embed_mat <- t(aligned[["slide_assay"]])
  embed_mat <- embed_mat[complete.cases(embed_mat), ]
  
  # Metadata: extract from colData and keep one row per patient
  coldata <- as.data.frame(colData(mae_emb))
  coldata$patient <- TCGAbarcode(rownames(coldata), participant = TRUE)
  meta <- coldata[!duplicated(coldata$patient), ]
  rownames(meta) <- meta$patient
  
  # Align metadata with embedding rows
  common <- intersect(rownames(embed_mat), rownames(meta))
  embed_meta <- embed_mat[common, ]
  meta_sub <- meta[common, ]
  
  return(list(embed = embed_meta, meta = meta_sub))
}


```


```{r}
# linkTCGA will match patients based on TCGA barcodes and keep only those
# with both molecular and slide data.
# Link TCGA
kidney_mae_emb <- linkTCGA(kidney_mae, catalog_kidney, redownload = T, parallel = T)
ovarian_mae_emb <- linkTCGA(OV_mae, catalog_ovar, redownload = T, parallel = T)
lung_mae_emb <- linkTCGA(lung_mae, catalog_lung, redownload = T, parallel = T)
coad_mae_emb <- linkTCGA(COAD_mae, catalog_coad, redownload = T, parallel = T)
read_mae_emb <- linkTCGA(READ_mae, catalog_rect, redownload = T, parallel = T)
# Get aligned embedding + metadata

kirc <- get_aligned_embedding_metadata(kidney_mae_emb)
ov <- get_aligned_embedding_metadata(ovarian_mae_emb)
read <- get_aligned_embedding_metadata(read_mae_emb)
coad <- get_aligned_embedding_metadata(coad_mae_emb)
lung <- get_aligned_embedding_metadata(lung_mae_emb)

```



* Check Patients

```{r}
# Number of patients after alignment
n_patients <- ncol(kirc[["slide_assay"]])
n_patients <- ncol(lung[["slide_assay"]])
cat("Patients with both embeddings and molecular data:", n_patients, "\n")
lapply(lung, dim)
```

# Metadata
```{r}

# Metadata 
embed_mat <- kirc$embed   # patients x features
meta_sub <- kirc$meta     # metadata aligned to patients

```



# Embedding structure

```{r}

# Compare most variable and least variable embedding features

# Compute variances for all features
all_vars <- apply(embed_mat, 2, var, na.rm = TRUE)

# Get top 20 highest variance and top 20 lowest variance
top20_high <- names(sort(all_vars, decreasing = TRUE))[1:20]
top20_low  <- names(sort(all_vars, decreasing = FALSE))[1:20]

# Values for bar plots
high_vals <- sort(all_vars, decreasing = TRUE)[1:20]
low_vals  <- sort(all_vars, decreasing = FALSE)[1:20]

# Set up plotting area (2 plots side by side)
par(mfrow = c(1, 2), mar = c(8, 4, 4, 2))

# Bar plot: most variable features
barplot(high_vals,
        names.arg = top20_high,
        las = 2,
        cex.names = 0.6,
        col = "firebrick3",
        ylab = "Variance",
        main = "Most variable embedding features\n(top 20 by variance)")

# Bar plot: least variable features
barplot(low_vals,
        names.arg = top20_low,
        las = 2,
        cex.names = 0.6,
        col = "steelblue",
        ylab = "Variance",
        main = "Least variable embedding features\n(top 20 lowest variance)")

# Reset plotting layout
par(mfrow = c(1, 1))



```


```{r}
# looking at the embedding values for the most variable features

# Prepare data for most variable features
high_df <- as.data.frame(embed_mat[, top20_high])
high_df$Patient <- rownames(high_df)
high_long <- pivot_longer(high_df, cols = -Patient,
                          names_to = "Feature", values_to = "Value")

# Prepare data for least variable features
low_df <- as.data.frame(embed_mat[, top20_low])
low_df$Patient <- rownames(low_df)
low_long <- pivot_longer(low_df, cols = -Patient,
                         names_to = "Feature", values_to = "Value")

# Boxplot for most variable
p1 <- ggplot(high_long, aes(x = Feature, y = Value)) +
  geom_boxplot(fill = "firebrick3", outlier.size = 0.5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(title = "Most variable features", y = "Feature value", x = "") +
  ylim(quantile(high_long$Value, c(0.01, 0.99), na.rm = TRUE))

# Boxplot for least variable
p2 <- ggplot(low_long, aes(x = Feature, y = Value)) +
  geom_boxplot(fill = "steelblue", outlier.size = 0.5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(title = "Least variable features", y = "Feature value", x = "") +
  ylim(quantile(low_long$Value, c(0.01, 0.99), na.rm = TRUE))

# Combine side by side

p1 + p2
```

## PCA
```{r}
# Compute PCA on aligned embeddings
pca_res <- prcomp(embed_mat, center = TRUE, scale. = TRUE)

# Build data frame for plotting
pca_df <- as.data.frame(pca_res$x[, 1:2])
colnames(pca_df) <- c("PC1", "PC2")
pca_df$Patient <- rownames(pca_df)
pca_df$Stage <- meta_sub[rownames(pca_df), "pathologic_stage"]

# Remove missing stage
pca_df <- pca_df[!is.na(pca_df$Stage), ]

# Keep stages with at least 3 patients (ellipse needs n>2)
stage_counts <- table(pca_df$Stage)
keep_stages <- names(stage_counts[stage_counts >= 3])
pca_df <- pca_df[pca_df$Stage %in% keep_stages, ]

# Variance explained
var_exp <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

# Plot with ellipses
ggplot(pca_df, aes(x = PC1, y = PC2, colour = Stage)) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = 2, size = 0.8) +
  scale_colour_viridis_d(option = "plasma") +
  labs(x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)"),
       title = "PCA of KIRC embeddings: coloured by pathologic stage",
       colour = "Stage") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5))


```


# Correlate embedding features and clinical variables

```{r}
clinical_vars <- data.frame(
  age = as.numeric(meta_sub$patient.age_at_initial_pathologic_diagnosis),
  sex = ifelse(meta_sub$gender == "MALE", 1, 0),
  pathologic_stage = as.numeric(as.factor(meta_sub$pathologic_stage)),
  T_stage = as.numeric(as.factor(meta_sub$pathology_T_stage)),
  N_stage = as.numeric(as.factor(meta_sub$pathology_N_stage)),
  M_stage = as.numeric(as.factor(meta_sub$pathology_M_stage)),
  row.names = rownames(meta_sub)
)
# Drop all-NA columns
clinical_vars <- clinical_vars[, colSums(!is.na(clinical_vars)) > 0, drop = FALSE]

# ----- 2. Embedding features (top 50 variable) -----
feature_vars <- apply(embed_mat, 2, var, na.rm = TRUE)
top50 <- order(feature_vars, decreasing = TRUE)[1:min(50, ncol(embed_mat))]
embed_top <- embed_mat[, top50]

# Remove zero‑variance embedding features
sd_feat <- apply(embed_top, 2, sd, na.rm = TRUE)
keep_feat <- which(!is.na(sd_feat) & sd_feat > 0)
embed_top <- embed_top[, keep_feat, drop = FALSE]

# ----- 3. Align patients -----
common <- intersect(rownames(embed_top), rownames(clinical_vars))
embed_aligned <- embed_top[common, , drop = FALSE]
clin_aligned <- clinical_vars[common, , drop = FALSE]

# Remove zero‑variance clinical variables
sd_clin <- apply(clin_aligned, 2, sd, na.rm = TRUE)
keep_clin <- which(!is.na(sd_clin) & sd_clin > 0)
clin_aligned <- clin_aligned[, keep_clin, drop = FALSE]

# If there are no variable features or no variable clinical variables, stop
if (ncol(embed_aligned) == 0) stop("No variable embedding features left.")
if (ncol(clin_aligned) == 0) stop("No variable clinical variables left.")

# ----- 4. Correlation matrix (features x clinical) -----
cor_mat <- cor(embed_aligned, clin_aligned, use = "pairwise.complete.obs")
# Remove any columns/rows that are all NA (should not happen after filtering)
cor_mat <- cor_mat[!apply(is.na(cor_mat), 1, all), , drop = FALSE]
cor_mat <- cor_mat[, !apply(is.na(cor_mat), 2, all), drop = FALSE]

# If matrix is empty, stop
if (nrow(cor_mat) == 0 || ncol(cor_mat) == 0) stop("Correlation matrix empty after cleaning.")


cluster_rows <- nrow(cor_mat) > 1
cluster_cols <- ncol(cor_mat) > 1

pheatmap(cor_mat,
         cluster_rows = cluster_rows,
         cluster_cols = cluster_cols,
         main = "Correlation: embedding features vs clinical variables",
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         angle_col = 45,
         fontsize_row = 8,
         fontsize_col = 10)
```


# Univariate associations with clinical features

```{r}

get_stage_coefficients <- function(embed_mat, meta_sub, fdr_thresh = 0.2) {
  stage_num <- as.numeric(as.factor(meta_sub$pathologic_stage))
  valid <- !is.na(stage_num)
  embed_valid <- embed_mat[valid, ]
  stage_valid <- stage_num[valid]
  
  coeff <- numeric(ncol(embed_valid))
  pval <- numeric(ncol(embed_valid))
  # here we'll run a simple lm for each single feature and extract the coefficient and p‑value for the feature term
  for (i in 1:ncol(embed_valid)) {
    fit <- lm(stage_valid ~ embed_valid[, i])
    coeff[i] <- coef(fit)[2]
    pval[i] <- summary(fit)$coefficients[2, 4]
  }
  names(coeff) <- colnames(embed_valid)
  names(pval) <- colnames(embed_valid)
  # using FDR to adjust, we'll set the filtering threshold to 0.2
  fdr <- p.adjust(pval, method = "fdr")
  sig <- which(fdr < fdr_thresh)
  # put everything in a df
  data.frame(
    Feature = names(coeff)[sig],
    Coefficient = coeff[sig],
    P_value = pval[sig],
    FDR = fdr[sig],
    stringsAsFactors = FALSE
  )
}
```


# run the linear models


```{r}
# Compute regression results
kirc_res <- get_stage_coefficients(kirc$embed, kirc$meta, fdr_thresh = 0.5)
coad_res <- get_stage_coefficients(coad$embed, coad$meta, fdr_thresh = 0.5)
lung_res <- get_stage_coefficients(lung$embed, lung$meta, fdr_thresh = 0.5)
ov_res <- get_stage_coefficients(ov$embed, ov$meta, fdr_thresh = 1)
read_res <- get_stage_coefficients(read$embed, read$meta, fdr_thresh = 0.5)
# Add cancer label
kirc_res$Cancer <- "KIRC"
coad_res$Cancer <- "COAD"
lung_res$Cancer <- "LUAD"
```

# Coeff Heatmap

## Visualize the top features associated with stage in each cancer, side by side, and then combined into one heatmap

```{r}

# Set parameters
n_top <- 30   # number of top features per cancer (by |coefficient|)

# KIRC: top features
kirc_top <- kirc_res[order(-abs(kirc_res$Coefficient)), ][1:n_top, ]
# Order from most positive to most negative for heatmap
kirc_top <- kirc_top[order(-kirc_top$Coefficient), ]
mat_kirc <- matrix(kirc_top$Coefficient, ncol = 1,
                   dimnames = list(kirc_top$Feature, "KIRC"))

# COAD: top features
coad_top <- coad_res[order(-abs(coad_res$Coefficient)), ][1:n_top, ]
coad_top <- coad_top[order(-coad_top$Coefficient), ]
mat_coad <- matrix(coad_top$Coefficient, ncol = 1,
                   dimnames = list(coad_top$Feature, "COAD"))


# Luad: top features
luad_top <- lung_res[order(-abs(lung_res$Coefficient)), ][1:n_top, ]
luad_top <- luad_top[order(-luad_top$Coefficient), ]
mat_luad <- matrix(luad_top$Coefficient, ncol = 1,
                   dimnames = list(luad_top$Feature, "LUAD"))

# Create individual heatmaps (pheatmap returns a grob when using `draw = FALSE`)

# Create heatmaps and capture the grob objects
p_kirc <- pheatmap(mat_kirc,
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                   fontsize_row = 8,
                   legend = TRUE,
                   legend_title = "Coefficient",
                   silent = TRUE) 

p_coad <- pheatmap(mat_coad,
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                   fontsize_row = 8,
                   legend = TRUE,
                   legend_title = "Coefficient",
                   silent = TRUE)


p_lung <- pheatmap(mat_luad,
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                   fontsize_row = 8,
                   legend = TRUE,
                   legend_title = "Coefficient",
                   silent = TRUE)

# Arrange side by side
library(gridExtra)
grid.arrange(p_kirc[[4]], p_coad[[4]], p_lung[[4]],  ncol = 3,
             top = "Top embedding features associated with pathologic stage (FDR < 0.2)")
```

* Now we combine the significant features from both cancers into one matrix for a single heatmap, sorted by the maximum coefficient across cancers (positive to negative)

```{r}
# Combine significant features from all three cancers
all_features <- unique(c(kirc_res$Feature, coad_res$Feature, lung_res$Feature))

# Build coefficient matrix (rows = features, columns = cancers)
coeff_matrix <- matrix(NA, nrow = length(all_features), ncol = 3,
                       dimnames = list(all_features, c("KIRC", "COAD", "LUAD")))

coeff_matrix[kirc_res$Feature, "KIRC"] <- kirc_res$Coefficient
coeff_matrix[coad_res$Feature, "COAD"] <- coad_res$Coefficient
coeff_matrix[lung_res$Feature, "LUAD"] <- lung_res$Coefficient
lung_res
# Remove rows that are all NA (should not happen, but safe)
coeff_matrix <- coeff_matrix[!apply(is.na(coeff_matrix), 1, all), ]

# Calculate mean coefficient across the three cancers (ignoring NAs)
mean_coef <- rowMeans(coeff_matrix, na.rm = TRUE)

# Keep top N features by absolute mean (strongest overall effect)
top_n <- 60
if (nrow(coeff_matrix) > top_n) {
  abs_mean <- abs(mean_coef)
  coeff_matrix <- coeff_matrix[order(abs_mean, decreasing = TRUE), ][1:top_n, ]
  # Recompute mean_coef after subsetting
  mean_coef <- rowMeans(coeff_matrix, na.rm = TRUE)
}

# Sort rows by mean coefficient (positive to negative)
coeff_matrix <- coeff_matrix[order(mean_coef, decreasing = TRUE), ]

# Heatmap
library(pheatmap)
pheatmap(coeff_matrix,
         cluster_rows = FALSE,          # keep our custom order
         cluster_cols = FALSE,
         na_col = "grey90",             # not significant in that cancer
         color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
         main = "Linear regression coefficients (stage ~ embedding feature)\nOrdered by mean coefficient (FDR < 0.2 in at least one cancer)",
         fontsize_row = 6,
         angle_col = 45,
         legend_title = "Coefficient β")

```

# Survival Cox univariate


### Prep Surv Data
```{r}
# ----- KIRC -----

kirc$meta$event <- as.numeric(kirc$meta$vital_status == 1)   # 1 = dead, 0 = alive
kirc$meta$time <- ifelse(kirc$meta$event == 1,
                         kirc$meta$days_to_death,
                         kirc$meta$patient.days_to_last_followup)
# Remove patients with missing time or event
kirc$meta <- kirc$meta[!is.na(kirc$meta$time) & !is.na(kirc$meta$event), ]

# ----- COAD -----
# Best columns: event = vital_status.x, death = days_to_death.x, follow = patient.days_to_last_followup
coad$meta$event <- as.numeric(coad$meta$vital_status.x == 1)   # 1 = dead, 0 = alive
coad$meta$time <- ifelse(coad$meta$event == 1,
                         coad$meta$days_to_death.x,
                         coad$meta$patient.days_to_last_followup)
# Remove rows with missing time or event
coad$meta <- coad$meta[!is.na(coad$meta$time) & !is.na(coad$meta$event), ]

# ----- LUAD -----

lung$meta$event <- as.numeric(lung$meta$vital_status == 1)   # 1 = dead, 0 = alive
lung$meta$time <- ifelse(lung$meta$event == 1,
                         lung$meta$days_to_death,
                         lung$meta$patient.days_to_last_followup)
# Remove rows with missing time or event
lung$meta <- lung$meta[!is.na(lung$meta$time) & !is.na(lung$meta$event), ]


# ----- OV -----

ov$meta$event <- as.numeric(ov$meta$vital_status == 1)   # 1 = dead, 0 = alive
ov$meta$time <- ifelse(ov$meta$event == 1,
                         ov$meta$days_to_death,
                         ov$meta$patient.days_to_last_followup)
# Remove rows with missing time or event
ov$meta <- ov$meta[!is.na(ov$meta$time) & !is.na(ov$meta$event), ]


# ----- READ -----

read$meta$event <- as.numeric(read$meta$vital_status == 1)   # 1 = dead, 0 = alive
read$meta$time <- ifelse(read$meta$event == 1,
                         read$meta$days_to_death,
                         read$meta$patient.days_to_last_followup)
# Remove rows with missing time or event
read$meta <- read$meta[!is.na(read$meta$time) & !is.na(read$meta$event), ]
```

```{r}
cat("KIRC: ", nrow(kirc$meta), " patients, events: ", sum(kirc$meta$event), "\n")
cat("COAD: ", nrow(coad$meta), " patients, events: ", sum(coad$meta$event), "\n")
cat("LUAD: ", nrow(lung$meta), " patients, events: ", sum(lung$meta$event), "\n")
cat("OV: ", nrow(ov$meta), " patients, events: ", sum(ov$meta$event), "\n")
cat("READ: ", nrow(read$meta), " patients, events: ", sum(read$meta$event), "\n")
```




```{r}
get_survival_coefficients <- function(embed_mat, meta_sub, time_col = "time", 
                                      event_col = "event", fdr_thresh = 0.2) {
  # embed_mat: patients x features
  # meta_sub: data frame with patient metadata (must contain time and event columns)
  # time_col: column name for survival time (days)
  # event_col: column name for event (1 = death, 0 = censored)
  
  # Align patients
  common <- intersect(rownames(embed_mat), rownames(meta_sub))
  embed <- embed_mat[common, ]
  meta <- meta_sub[common, ]
  
  # Extract time and event
  time <- meta[[time_col]]
  event <- meta[[event_col]]
  
  # Remove patients with missing time or event
  valid <- complete.cases(time, event)
  embed <- embed[valid, ]
  time <- time[valid]
  event <- event[valid]
  
  # Build survival object
  surv_obj <- Surv(time, event)
  
  # Loop over each embedding feature
  coeff <- numeric(ncol(embed))
  pval <- numeric(ncol(embed))
  
  for (i in 1:ncol(embed)) {
    fit <- coxph(surv_obj ~ embed[, i])
    coeff[i] <- coef(fit)[1]           # log hazard ratio
    pval[i] <- summary(fit)$coefficients[1, 5]  # p-value for the coefficient
  }
  
  names(coeff) <- colnames(embed)
  names(pval) <- colnames(embed)
  
  fdr <- p.adjust(pval, method = "fdr")
  sig <- which(fdr < fdr_thresh)
  
  data.frame(
    Feature = names(coeff)[sig],
    LogHR = coeff[sig],
    P_value = pval[sig],
    FDR = fdr[sig],
    stringsAsFactors = FALSE
  )
}
```


```{r}

# For KIRC
kirc_surv <- get_survival_coefficients(kirc$embed, kirc$meta,
                                       time_col = "time",
                                       event_col = "event",
                                       fdr_thresh = 0.5)

# For COAD
coad_surv <- get_survival_coefficients(coad$embed, coad$meta,
                                       time_col = "time",
                                       event_col = "event",
                                       fdr_thresh = 0.2)

# For LUAD
lung_surv <- get_survival_coefficients(lung$embed, lung$meta,
                                       time_col = "time",
                                       event_col = "event",
                                       fdr_thresh = 0.2)

# For OV
ov_surv <- get_survival_coefficients(ov$embed, ov$meta,
                                       time_col = "time",
                                       event_col = "event",
                                       fdr_thresh = 0.2)

# For Read
read_surv <- get_survival_coefficients(read$embed, read$meta,
                                       time_col = "time",
                                       event_col = "event",
                                       fdr_thresh = 0.8)


# Add cancer labels
kirc_surv$Cancer <- "KIRC"
coad_surv$Cancer <- "COAD"
lung_surv$Cancer <- "LUAD"
ov_surv$Cancer <- "OV"
read_surv$Cancer <- "READ"

```


```{r}
# List all cancer survival results
all_surv <- list(
  KIRC = kirc_surv,
  COAD = coad_surv,
  LUAD = lung_surv,
  OV   = ov_surv,
  READ = read_surv
)

# ------------------------------
# 1. Select top 20 features per cancer (by absolute LogHR)
# ------------------------------
n_top_per_cancer <- 30

top_features_per_cancer <- lapply(all_surv, function(df) {
  df %>%
    arrange(desc(abs(LogHR))) %>%
    slice_head(n = n_top_per_cancer) %>%
    pull(Feature)
})

# Union of all top features
all_features <- unique(unlist(top_features_per_cancer))

# ------------------------------
# 2. Build coefficient matrix (only rows for those features)
# ------------------------------
coeff_matrix <- matrix(NA, nrow = length(all_features), ncol = length(all_surv),
                       dimnames = list(all_features, names(all_surv)))

for (cancer in names(all_surv)) {
  df <- all_surv[[cancer]]
  # Only assign for features that exist in the matrix rows
  common <- intersect(df$Feature, rownames(coeff_matrix))
  if (length(common) > 0) {
    coeff_matrix[common, cancer] <- df$LogHR[match(common, df$Feature)]
  }
}

# Remove any all‑NA rows (shouldn't happen, but safe)
coeff_matrix <- coeff_matrix[!apply(is.na(coeff_matrix), 1, all), ]

# ------------------------------
# 3. Order rows by mean LogHR (positive to negative)
# ------------------------------
mean_coef <- rowMeans(coeff_matrix, na.rm = TRUE)
coeff_matrix <- coeff_matrix[order(mean_coef, decreasing = TRUE), ]

# ------------------------------
# 4. Heatmap
# ------------------------------
pheatmap(coeff_matrix,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         na_col = "grey90",
         color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
         main = paste0("Log Hazard Ratios (survival ~ embedding feature)\nTop ", n_top_per_cancer, " features per cancer (union)"),
         fontsize_row = 6,
         angle_col = 45,
         legend_title = "Log HR")
```



# Compute pathway scores (GSVA) and correlate with embeddings

1. We build pathway_scores from the RNA‑seq data via GSVA – patients × pathways.

2. We have the embedding matrix – patients × features.

3. We correlate features with pathways for the same patients.

4. We take the intersection of patient IDs to ensure both matrices have identical rows.

5. We subset both matrices to those common patients.

6. Finally, we compute cor(embed_aligned, pathway_aligned) – each row of embeds corresponds to the same patient as the corresponding row of pathways.



```{r}
get_gsva_pathway <- function(mae_emb, rna_assay_name, aligned_list, res, 
                             top_n = NULL, min_n = 30) {
  # 1. Get Hallmark gene sets
  hallmark <- msigdbr(species = "Homo sapiens", collection = "H") %>%
    split(x = .$gene_symbol, f = .$gs_name)
  
  # 2. Extract RNA and align to the patients in aligned_list$embed
  rna_se <- mae_emb[[rna_assay_name]]
  rna_mat <- assay(rna_se)
  sm <- sampleMap(mae_emb)
  sm_rna <- sm[sm$assay == rna_assay_name, ]
  patients <- rownames(aligned_list$embed)   # already standardised
  
  rna_patient <- matrix(NA, nrow = nrow(rna_mat), ncol = length(patients),
                        dimnames = list(rownames(rna_mat), patients))
  for (pat in patients) {
    cols <- sm_rna$colname[sm_rna$primary == pat]
    if (length(cols) > 0) {
      rna_patient[, pat] <- rowMeans(rna_mat[, cols, drop = FALSE], na.rm = TRUE)
    }
  }
  # Drop genes that are all NA
  keep <- rowSums(!is.na(rna_patient)) > 0
  rna_patient <- rna_patient[keep, , drop = FALSE]
  
  # 3. GSVA
  ssgsea_param <- ssgseaParam(exprData = rna_patient, geneSets = hallmark,
                              alpha = 0.25, normalize = TRUE)
  gsva_scores <- gsva(ssgsea_param, verbose = FALSE)
  pathway_scores <- t(gsva_scores)
  
  # 4. Standardise pathway score patient IDs (to match embedding IDs)
  rownames(pathway_scores) <- TCGAbarcode(rownames(pathway_scores), participant = TRUE)
  
  # 5. Align patients – `aligned_list$embed` already has standardised IDs
  common <- intersect(rownames(aligned_list$embed), rownames(pathway_scores))
  if (length(common) < 3) {
    warning("Fewer than 3 common patients – returning empty.")
    return(data.frame())
  }
  embed_aligned <- aligned_list$embed[common, , drop = FALSE]
  pathway_aligned <- pathway_scores[common, , drop = FALSE]
  
  # 6. Remove sparse pathways (too many NAs)
  n_valid <- colSums(!is.na(pathway_aligned))
  pathway_aligned <- pathway_aligned[, n_valid >= min_n, drop = FALSE]
  
  # 7. Remove constant pathways (zero variance)
  const_path <- apply(pathway_aligned, 2, function(x) length(unique(x)) == 1)
  pathway_aligned <- pathway_aligned[, !const_path, drop = FALSE]
  
  # 8. Remove zero‑variance embedding features
  feat_var <- apply(embed_aligned, 2, var, na.rm = TRUE)
  embed_aligned <- embed_aligned[, feat_var > 1e-6, drop = FALSE]
  
  # 9. Spearman correlation
  cor_embed_pathway <- cor(embed_aligned, pathway_aligned, 
                           method = "spearman", use = "pairwise.complete.obs")
  
  # 10. Blank correlations with effective n < min_n
  n_pairs <- crossprod(!is.na(embed_aligned), !is.na(pathway_aligned))
  cor_embed_pathway[n_pairs < min_n] <- NA
  
  # 11. Select top features from regression results
  coef_col <- if ("Coefficient" %in% colnames(res)) "Coefficient" else "LogHR"
  if (!is.numeric(res[[coef_col]])) res[[coef_col]] <- as.numeric(res[[coef_col]])
  
  selected_features <- if (!is.null(top_n) && top_n < nrow(res)) {
    head(res[order(-abs(res[[coef_col]])), "Feature"], top_n)
  } else {
    res$Feature
  }
  
  # 12. Subset correlation matrix
  pathway_cor_subset <- cor_embed_pathway[selected_features, , drop = FALSE]
  
  # 13. For each feature, get top 3 positive and negative pathways
  pathway_summary <- list()
  for (feat in selected_features) {
    cors <- pathway_cor_subset[feat, ]
    cors <- cors[!is.na(cors)]
    if (length(cors) == 0) next
    pos <- sort(cors, decreasing = TRUE)[1:min(3, length(cors))]
    neg <- sort(cors, decreasing = FALSE)[1:min(3, length(cors))]
    pathway_summary[[feat]] <- data.frame(
      Embedding_Feature = feat,
      Direction = rep(c("Positive", "Negative"), each = length(pos)),
      Pathway = c(names(pos), names(neg)),
      Correlation = c(pos, neg),
      stringsAsFactors = FALSE
    )
  }
  
  pathway_df <- do.call(rbind, pathway_summary)
  rownames(pathway_df) <- NULL
  
  return(pathway_df)
}

```


* Note here: For a given embedding feature (e.g., slide_feat_144) and a given pathway (e.g., HALLMARK_APICAL_JUNCTION), we compute the Spearman correlation between the vector of embedding values across patients and the vector of pathway scores across the same patients.


* To avoid perfect corr and issues with missingness, we will filter so the data has at least min of 30 non‑NA patients to ensure reliability.


```{r}

# KIRC – use kirc_surv

pathway_kirc <- get_gsva_pathway(
  mae_emb = kidney_mae_emb,
  rna_assay_name = "KIRC_RNASeq2GeneNorm-20160128",
  aligned_list = kirc,
  res = kirc_surv,  
  top_n = 20,
  min_n = 30
)

# COAD

pathway_coad <- get_gsva_pathway(
  mae_emb = coad_mae_emb,
  rna_assay_name = "COAD_RNASeq2GeneNorm-20160128",
  aligned_list = coad,
  res = coad_surv,
  top_n = 30,
      min_n = 30

)

# LUAD
pathway_luad <- get_gsva_pathway(
  mae_emb = lung_mae_emb,
  rna_assay_name = "LUAD_RNASeq2Gene-20160128",
  aligned_list = lung,
  res = lung_surv,  
  top_n = 30,
  min_n = 30
)

# OV

pathway_ov <- get_gsva_pathway(
  mae_emb = ovarian_mae_emb,
  rna_assay_name = "OV_RNASeq2GeneNorm-20160128",
  aligned_list = ov,
  res = ov_surv,    
  top_n = 30,
  min_n = 30
)

# READ
pathway_read <- get_gsva_pathway(
  mae_emb = read_mae_emb,
  rna_assay_name = "READ_RNASeq2GeneNorm-20160128",
  aligned_list = read,
  res = read_surv, 
  top_n = 30,
  min_n = 30
)
```

## Check overlapping
```{r}
# Get feature lists
surv_features <- rownames(coeff_matrix)  # from your survival heatmap
pathway_features <- unique(pathway_ov$Embedding_Feature)

# Overlap
intersect(surv_features, pathway_features)

# Features in survival but missing from pathway
setdiff(surv_features, pathway_features)
```



# Check Missingness
```{r}


# Function to summarise missingness for a given assay
assay_missingness <- function(mae_emb, assay_name) {
  se <- mae_emb[[assay_name]]
  mat <- assay(se)
  
  # Overall missingness
  total_cells <- prod(dim(mat))
  missing_cells <- sum(is.na(mat))
  missing_pct <- 100 * missing_cells / total_cells
  
  cat(sprintf("\n[%s] Total cells: %d, Missing: %d (%.2f%%)\n", 
              assay_name, total_cells, missing_cells, missing_pct))
  
  # Per‑sample missingness
  sample_missing <- colSums(is.na(mat))
  df_sample <- data.frame(Sample = colnames(mat), Missing = sample_missing)
  
  p1 <- ggplot(df_sample, aes(x = reorder(Sample, Missing), y = Missing)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    labs(title = paste("Missing values per sample:", assay_name),
         x = "Sample", y = "Number of missing genes") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 6))
  
  # Per‑gene missingness
  gene_missing <- rowSums(is.na(mat))
  p2 <- ggplot(data.frame(Missing = gene_missing), aes(x = Missing)) +
    geom_histogram(bins = 30, fill = "firebrick3", alpha = 0.7) +
    labs(title = paste("Missing values per gene:", assay_name),
         x = "Number of samples with missing value", y = "Frequency") +
    theme_minimal()
  
  return(list(overall = c(total_cells, missing_cells, missing_pct),
              sample_plot = p1, gene_plot = p2))
}

out_kirc <- assay_missingness(kidney_mae_emb, "KIRC_RNASeq2GeneNorm-20160128")
out_kirc$sample_plot
out_kirc$gene_plot


out_ov <- assay_missingness(ovarian_mae_emb, "OV_RNASeq2GeneNorm-20160128")
out_ov$sample_plot
out_ov$gene_plot
```


# GSVA Embedding corr table

```{r}
pathway_list <- list(
  KIRC = pathway_kirc,
  COAD = pathway_coad,
  LUAD = pathway_luad,
  OV   = pathway_ov,
  READ = pathway_read
)

summary_list <- list()

for (cancer in names(pathway_list)) {
  df <- pathway_list[[cancer]]
  if (is.null(df) || nrow(df) == 0) next
  
  # For each feature, pick the strongest association (highest |Correlation|)
  top_per_feature <- df %>%
    group_by(Embedding_Feature) %>%
    slice_max(order_by = abs(Correlation), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(abs(Correlation))) %>%
    slice_head(n = 5)   # keep top 5 features per cancer
  
  if (nrow(top_per_feature) > 0) {
    # Add LogHR from survival matrix
    top_per_feature$LogHR <- sapply(top_per_feature$Embedding_Feature, function(f) {
      if (f %in% rownames(coeff_matrix)) {
        coeff_matrix[f, cancer]
      } else {
        NA
      }
    })
    top_per_feature$Cancer <- cancer
    summary_list[[cancer]] <- top_per_feature
  }
}

# Combine and reorder columns
final_table <- bind_rows(summary_list) %>%
  select(Cancer, Embedding_Feature, LogHR, Pathway, Correlation, Direction)

# Optional: filter to |Correlation| > 0.2
final_table <- final_table %>% filter(abs(Correlation) > 0.2)


```

```{r}
# Prepare data
sign_embed_data <- final_table %>%
  mutate(
    Cancer = factor(Cancer, levels = c("KIRC", "COAD", "LUAD", "OV", "READ")),
    Label = paste0(Embedding_Feature, "\n", Pathway)
  )

# Scatter plot
ggplot(sign_embed_data, aes(x = LogHR, y = Correlation, color = Cancer)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_text_repel(aes(label = Label), size = 3, max.overlaps = 15) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Survival-associated embeddings and their pathway correlations",
    x = "Log Hazard Ratio (survival)",
    y = "Spearman correlation with Hallmark pathway",
    color = "Cancer"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```



# Full Table

```{r}
# Assume final_table is your existing data frame

final_table_with_links <- final_table %>%
  mutate(
    # Add clickable link
    # Add a placeholder interpretation (you can replace these with your own text)
    Interpretation = case_when(
      Pathway == "HALLMARK_APICAL_JUNCTION" ~ "Cell–cell adhesion; may reflect epithelial disorganisation.",
      Pathway == "HALLMARK_IL6_JAK_STAT3_SIGNALING" ~ "Inflammatory signalling; often pro‑tumorigenic.",
      Pathway == "HALLMARK_HEME_METABOLISM" ~ "Erythroid/metabolic pathway; suggests metabolic reprogramming.",
      Pathway == "HALLMARK_ADIPOGENESIS" ~ "Adipocyte differentiation; may reflect tumour‑stroma interaction.",
      Pathway == "HALLMARK_PEROXISOME" ~ "Lipid metabolism; altered in aggressive tumours.",
      Pathway == "HALLMARK_ESTROGEN_RESPONSE_LATE" ~ "Hormone signalling; often relevant in breast/colon cancer.",
      Pathway == "HALLMARK_MTORC1_SIGNALING" ~ "Nutrient sensing and growth; high activity linked to poor prognosis.",
      Pathway == "HALLMARK_G2M_CHECKPOINT" ~ "Cell cycle control; high activity often indicates proliferation.",
      Pathway == "HALLMARK_MITOTIC_SPINDLE" ~ "Mitotic progression; reflects proliferative capacity.",
      Pathway == "HALLMARK_ANDROGEN_RESPONSE" ~ "Androgen signalling; relevant in ovarian and other cancers.",
      Pathway == "HALLMARK_CHOLESTEROL_HOMEOSTASIS" ~ "Cholesterol metabolism; altered in cancer.",
      Pathway == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" ~ "Inflammatory/NF‑κB signalling; linked to progression.",
      Pathway == "HALLMARK_PROTEIN_SECRETION" ~ "Secretory pathway; increased in aggressive tumours.",
      Pathway == "HALLMARK_UV_RESPONSE_DN" ~ "DNA damage response; down‑regulated in some tumours.",
      TRUE ~ "Review pathway definition."
    )
  ) %>%
  select(Cancer, Embedding_Feature, LogHR, Pathway, Correlation, Direction)



library(kableExtra)
final_table_with_links %>%
  kbl(
    caption = "Top survival‑associated embeddings and their pathway correlations",
    digits = 3,
    escape = FALSE
  ) %>%
  kable_classic(full_width = FALSE)

```

