---
title: "embed_analysis"
format: html
editor: visual
---

## Libraries

```{r, message=FALSE}
library(rmarkdown)
library(knitr)
library(psych)
#BiocManager::install("waldronlab/imageFeatureTCGA")
packageVersion("imageFeatureTCGA")
library(imageFeatureTCGA)
library(curatedTCGAData)
library(SummarizedExperiment)
library(dplyr)
library(TCGAutils)
library(ggplot2)
library(pheatmap)
library(GSVA)
library(msigdbr)
library(tidyr)
library(patchwork)

```


# Load slide data catalog (Kidney Renal Cancer)

```{r load embeddings}
# Kidney
catalog_kidney <- getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-KIRC")


# Ovarian
catalog_lung <- getCatalog("provgigapath") |> 
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-LUAD")


# Colorectal
catalog_coad <- getCatalog("provgigapath") |> 
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-COAD")



```


# Read Cancer assays with curatedTCGA

```{r load_rnaseq, message=FALSE}

# Load Kidney assays
kidney_mae <- curatedTCGAData(
    diseaseCode = "KIRC",
    assays =  c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
    version = "2.1.1",
    dry.run = FALSE
)

lung_mae <- curatedTCGAData(
  diseaseCode = "LUAD",
  assays = c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
  version = "2.1.1",
  dry.run = FALSE
)

COAD_mae <- curatedTCGAData(
  diseaseCode = "COAD",
  assays = c("RNASeq2GeneNorm", "RPPAArray", "mRNAArray"),
  version = "2.1.1",
  dry.run = FALSE
)

```


# Align embeddings

### Helper Function

```{r}
get_aligned_embedding_metadata <- function(mae_emb) {
  # mae_emb is output from linkTCGA (MultiAssayExperiment with slide_assay)
  # Align patients across assays and return embedding matrix (patients x features) and metadata
  
  # Add row names to slide_assay if not already
  if (is.null(rownames(mae_emb[["slide_assay"]]))) {
    rownames(mae_emb[["slide_assay"]]) <- paste0("slide_feat_", seq_len(nrow(mae_emb[["slide_assay"]])))
  }
  
  # Get patients with slide data
  slide_sm <- sampleMap(mae_emb)[sampleMap(mae_emb)$assay == "slide_assay", ]
  patients <- unique(slide_sm$primary)
  
  # Build patient-level matrices for each assay
  aligned <- list()
  for (nm in names(experiments(mae_emb))) {
    se <- experiments(mae_emb)[[nm]]
    mat <- assay(se)
    sm <- sampleMap(mae_emb)[sampleMap(mae_emb)$assay == nm, ]
    col_patients <- sm$primary[match(colnames(mat), sm$colname)]
    patient_mat <- matrix(NA, nrow = nrow(mat), ncol = length(patients),
                          dimnames = list(rownames(mat), patients))
    for (pat in patients) {
      cols <- which(col_patients == pat)
      if (length(cols) == 1) patient_mat[, pat] <- mat[, cols]
      else if (length(cols) > 1) patient_mat[, pat] <- rowMeans(mat[, cols, drop = FALSE], na.rm = TRUE)
    }
    if (!is.numeric(patient_mat)) {
      patient_mat <- matrix(as.numeric(patient_mat), nrow = nrow(patient_mat),
                            dimnames = dimnames(patient_mat))
    }
    keep <- rowSums(!is.na(patient_mat)) > 0
    aligned[[nm]] <- patient_mat[keep, , drop = FALSE]
  }
  
  # Embedding matrix (features x patients) -> transpose to patients x features
  embed_mat <- t(aligned[["slide_assay"]])
  embed_mat <- embed_mat[complete.cases(embed_mat), ]
  
  # Metadata: extract from colData and keep one row per patient
  coldata <- as.data.frame(colData(mae_emb))
  coldata$patient <- TCGAbarcode(rownames(coldata), participant = TRUE)
  meta <- coldata[!duplicated(coldata$patient), ]
  rownames(meta) <- meta$patient
  
  # Align metadata with embedding rows
  common <- intersect(rownames(embed_mat), rownames(meta))
  embed_meta <- embed_mat[common, ]
  meta_sub <- meta[common, ]
  
  return(list(embed = embed_meta, meta = meta_sub))
}


```


```{r}
# linkTCGA will match patients based on TCGA barcodes and keep only those
# with both molecular and slide data.
# Link TCGA
kidney_mae_emb <- linkTCGA(kidney_mae, catalog_kidney, redownload = T, parallel = T)
#ovarian_mae_emb <- linkTCGA(ov_mae, catalog_ovarian, redownload = T, parallel = T)
lung_mae_emb <- linkTCGA(lung_mae, catalog_lung, redownload = T, parallel = T)
coad_mae_emb <- linkTCGA(COAD_mae, catalog_coad, redownload = T, parallel = T)
# Get aligned embedding + metadata

kirc <- get_aligned_embedding_metadata(kidney_mae_emb)
#ov <- get_aligned_embedding_metadata(ovarian_mae_emb)
coad <- get_aligned_embedding_metadata(coad_mae_emb)
lung <- get_aligned_embedding_metadata(lung_mae_emb)

```



* Check Patients

```{r}
# Number of patients after alignment
n_patients <- ncol(kirc[["slide_assay"]])
n_patients <- ncol(lung[["slide_assay"]])
cat("Patients with both embeddings and molecular data:", n_patients, "\n")
lapply(lung, dim)
```

# Metadata
```{r}

# Metadata 
embed_mat <- kirc$embed   # patients x features
meta_sub <- kirc$meta     # metadata aligned to patients

```



# Embedding structure

```{r}

# Compare most variable and least variable embedding features

# Compute variances for all features
all_vars <- apply(embed_mat, 2, var, na.rm = TRUE)

# Get top 20 highest variance and top 20 lowest variance
top20_high <- names(sort(all_vars, decreasing = TRUE))[1:20]
top20_low  <- names(sort(all_vars, decreasing = FALSE))[1:20]

# Values for bar plots
high_vals <- sort(all_vars, decreasing = TRUE)[1:20]
low_vals  <- sort(all_vars, decreasing = FALSE)[1:20]

# Set up plotting area (2 plots side by side)
par(mfrow = c(1, 2), mar = c(8, 4, 4, 2))

# Bar plot: most variable features
barplot(high_vals,
        names.arg = top20_high,
        las = 2,
        cex.names = 0.6,
        col = "firebrick3",
        ylab = "Variance",
        main = "Most variable embedding features\n(top 20 by variance)")

# Bar plot: least variable features
barplot(low_vals,
        names.arg = top20_low,
        las = 2,
        cex.names = 0.6,
        col = "steelblue",
        ylab = "Variance",
        main = "Least variable embedding features\n(top 20 lowest variance)")

# Reset plotting layout
par(mfrow = c(1, 1))



```


```{r}
# looking at the embedding values for the most variable features

# Prepare data for most variable features
high_df <- as.data.frame(embed_mat[, top20_high])
high_df$Patient <- rownames(high_df)
high_long <- pivot_longer(high_df, cols = -Patient,
                          names_to = "Feature", values_to = "Value")

# Prepare data for least variable features
low_df <- as.data.frame(embed_mat[, top20_low])
low_df$Patient <- rownames(low_df)
low_long <- pivot_longer(low_df, cols = -Patient,
                         names_to = "Feature", values_to = "Value")

# Boxplot for most variable
p1 <- ggplot(high_long, aes(x = Feature, y = Value)) +
  geom_boxplot(fill = "firebrick3", outlier.size = 0.5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(title = "Most variable features", y = "Feature value", x = "") +
  ylim(quantile(high_long$Value, c(0.01, 0.99), na.rm = TRUE))

# Boxplot for least variable
p2 <- ggplot(low_long, aes(x = Feature, y = Value)) +
  geom_boxplot(fill = "steelblue", outlier.size = 0.5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(title = "Least variable features", y = "Feature value", x = "") +
  ylim(quantile(low_long$Value, c(0.01, 0.99), na.rm = TRUE))

# Combine side by side

p1 + p2
```

## PCA
```{r}
# Compute PCA on aligned embeddings
pca_res <- prcomp(embed_mat, center = TRUE, scale. = TRUE)

# Build data frame for plotting
pca_df <- as.data.frame(pca_res$x[, 1:2])
colnames(pca_df) <- c("PC1", "PC2")
pca_df$Patient <- rownames(pca_df)
pca_df$Stage <- meta_sub[rownames(pca_df), "pathologic_stage"]

# Remove missing stage
pca_df <- pca_df[!is.na(pca_df$Stage), ]

# Keep stages with at least 3 patients (ellipse needs n>2)
stage_counts <- table(pca_df$Stage)
keep_stages <- names(stage_counts[stage_counts >= 3])
pca_df <- pca_df[pca_df$Stage %in% keep_stages, ]

# Variance explained
var_exp <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

# Plot with ellipses
ggplot(pca_df, aes(x = PC1, y = PC2, colour = Stage)) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = 2, size = 0.8) +
  scale_colour_viridis_d(option = "plasma") +
  labs(x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)"),
       title = "PCA of KIRC embeddings: coloured by pathologic stage",
       colour = "Stage") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5))


```


# Correlate embedding features and clinical variables

```{r}
clinical_vars <- data.frame(
  age = as.numeric(meta_sub$patient.age_at_initial_pathologic_diagnosis),
  sex = ifelse(meta_sub$gender == "MALE", 1, 0),
  pathologic_stage = as.numeric(as.factor(meta_sub$pathologic_stage)),
  T_stage = as.numeric(as.factor(meta_sub$pathology_T_stage)),
  N_stage = as.numeric(as.factor(meta_sub$pathology_N_stage)),
  M_stage = as.numeric(as.factor(meta_sub$pathology_M_stage)),
  row.names = rownames(meta_sub)
)
# Drop all-NA columns
clinical_vars <- clinical_vars[, colSums(!is.na(clinical_vars)) > 0, drop = FALSE]

# ----- 2. Embedding features (top 50 variable) -----
feature_vars <- apply(embed_mat, 2, var, na.rm = TRUE)
top50 <- order(feature_vars, decreasing = TRUE)[1:min(50, ncol(embed_mat))]
embed_top <- embed_mat[, top50]

# Remove zero‑variance embedding features
sd_feat <- apply(embed_top, 2, sd, na.rm = TRUE)
keep_feat <- which(!is.na(sd_feat) & sd_feat > 0)
embed_top <- embed_top[, keep_feat, drop = FALSE]

# ----- 3. Align patients -----
common <- intersect(rownames(embed_top), rownames(clinical_vars))
embed_aligned <- embed_top[common, , drop = FALSE]
clin_aligned <- clinical_vars[common, , drop = FALSE]

# Remove zero‑variance clinical variables
sd_clin <- apply(clin_aligned, 2, sd, na.rm = TRUE)
keep_clin <- which(!is.na(sd_clin) & sd_clin > 0)
clin_aligned <- clin_aligned[, keep_clin, drop = FALSE]

# If there are no variable features or no variable clinical variables, stop
if (ncol(embed_aligned) == 0) stop("No variable embedding features left.")
if (ncol(clin_aligned) == 0) stop("No variable clinical variables left.")

# ----- 4. Correlation matrix (features x clinical) -----
cor_mat <- cor(embed_aligned, clin_aligned, use = "pairwise.complete.obs")
# Remove any columns/rows that are all NA (should not happen after filtering)
cor_mat <- cor_mat[!apply(is.na(cor_mat), 1, all), , drop = FALSE]
cor_mat <- cor_mat[, !apply(is.na(cor_mat), 2, all), drop = FALSE]

# If matrix is empty, stop
if (nrow(cor_mat) == 0 || ncol(cor_mat) == 0) stop("Correlation matrix empty after cleaning.")


cluster_rows <- nrow(cor_mat) > 1
cluster_cols <- ncol(cor_mat) > 1

pheatmap(cor_mat,
         cluster_rows = cluster_rows,
         cluster_cols = cluster_cols,
         main = "Correlation: embedding features vs clinical variables",
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         angle_col = 45,
         fontsize_row = 8,
         fontsize_col = 10)
```


# Univariate associations with clinical features

```{r}

get_stage_coefficients <- function(embed_mat, meta_sub, fdr_thresh = 0.2) {
  stage_num <- as.numeric(as.factor(meta_sub$pathologic_stage))
  valid <- !is.na(stage_num)
  embed_valid <- embed_mat[valid, ]
  stage_valid <- stage_num[valid]
  
  coeff <- numeric(ncol(embed_valid))
  pval <- numeric(ncol(embed_valid))
  # here we'll run a simple lm for each single feature and extract the coefficient and p‑value for the feature term
  for (i in 1:ncol(embed_valid)) {
    fit <- lm(stage_valid ~ embed_valid[, i])
    coeff[i] <- coef(fit)[2]
    pval[i] <- summary(fit)$coefficients[2, 4]
  }
  names(coeff) <- colnames(embed_valid)
  names(pval) <- colnames(embed_valid)
  # using FDR to adjust, we'll set the filtering threshold to 0.2
  fdr <- p.adjust(pval, method = "fdr")
  sig <- which(fdr < fdr_thresh)
  # put everything in a df
  data.frame(
    Feature = names(coeff)[sig],
    Coefficient = coeff[sig],
    P_value = pval[sig],
    FDR = fdr[sig],
    stringsAsFactors = FALSE
  )
}
```


# run the linear models


```{r}
# Compute regression results
kirc_res <- get_stage_coefficients(kirc$embed, kirc$meta, fdr_thresh = 0.5)
coad_res <- get_stage_coefficients(coad$embed, coad$meta, fdr_thresh = 0.5)
lung_res <- get_stage_coefficients(lung$embed, lung$meta, fdr_thresh = 0.5)
# Add cancer label
kirc_res$Cancer <- "KIRC"
coad_res$Cancer <- "COAD"
lung_res$Cancer <- "LUAD"
```

# Coeff Heatmap

## Visualize the top features associated with stage in each cancer, side by side, and then combined into one heatmap

```{r}

# Set parameters
n_top <- 30   # number of top features per cancer (by |coefficient|)

# KIRC: top features
kirc_top <- kirc_res[order(-abs(kirc_res$Coefficient)), ][1:n_top, ]
# Order from most positive to most negative for heatmap
kirc_top <- kirc_top[order(-kirc_top$Coefficient), ]
mat_kirc <- matrix(kirc_top$Coefficient, ncol = 1,
                   dimnames = list(kirc_top$Feature, "KIRC"))

# COAD: top features
coad_top <- coad_res[order(-abs(coad_res$Coefficient)), ][1:n_top, ]
coad_top <- coad_top[order(-coad_top$Coefficient), ]
mat_coad <- matrix(coad_top$Coefficient, ncol = 1,
                   dimnames = list(coad_top$Feature, "COAD"))


# Luad: top features
luad_top <- lung_res[order(-abs(lung_res$Coefficient)), ][1:n_top, ]
luad_top <- luad_top[order(-luad_top$Coefficient), ]
mat_luad <- matrix(luad_top$Coefficient, ncol = 1,
                   dimnames = list(luad_top$Feature, "LUAD"))

# Create individual heatmaps (pheatmap returns a grob when using `draw = FALSE`)

# Create heatmaps and capture the grob objects
p_kirc <- pheatmap(mat_kirc,
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                   fontsize_row = 8,
                   legend = TRUE,
                   legend_title = "Coefficient",
                   silent = TRUE) 

p_coad <- pheatmap(mat_coad,
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                   fontsize_row = 8,
                   legend = TRUE,
                   legend_title = "Coefficient",
                   silent = TRUE)


p_lung <- pheatmap(mat_luad,
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                   fontsize_row = 8,
                   legend = TRUE,
                   legend_title = "Coefficient",
                   silent = TRUE)

# Arrange side by side
library(gridExtra)
grid.arrange(p_kirc[[4]], p_coad[[4]], p_lung[[4]],  ncol = 3,
             top = "Top embedding features associated with pathologic stage (FDR < 0.2)")
```

* Now we combine the significant features from both cancers into one matrix for a single heatmap, sorted by the maximum coefficient across cancers (positive to negative)

```{r}
# Combine significant features from all three cancers
all_features <- unique(c(kirc_res$Feature, coad_res$Feature, lung_res$Feature))

# Build coefficient matrix (rows = features, columns = cancers)
coeff_matrix <- matrix(NA, nrow = length(all_features), ncol = 3,
                       dimnames = list(all_features, c("KIRC", "COAD", "LUAD")))

coeff_matrix[kirc_res$Feature, "KIRC"] <- kirc_res$Coefficient
coeff_matrix[coad_res$Feature, "COAD"] <- coad_res$Coefficient
coeff_matrix[lung_res$Feature, "LUAD"] <- lung_res$Coefficient
lung_res
# Remove rows that are all NA (should not happen, but safe)
coeff_matrix <- coeff_matrix[!apply(is.na(coeff_matrix), 1, all), ]

# Calculate mean coefficient across the three cancers (ignoring NAs)
mean_coef <- rowMeans(coeff_matrix, na.rm = TRUE)

# Keep top N features by absolute mean (strongest overall effect)
top_n <- 60
if (nrow(coeff_matrix) > top_n) {
  abs_mean <- abs(mean_coef)
  coeff_matrix <- coeff_matrix[order(abs_mean, decreasing = TRUE), ][1:top_n, ]
  # Recompute mean_coef after subsetting
  mean_coef <- rowMeans(coeff_matrix, na.rm = TRUE)
}

# Sort rows by mean coefficient (positive to negative)
coeff_matrix <- coeff_matrix[order(mean_coef, decreasing = TRUE), ]

# Heatmap
library(pheatmap)
pheatmap(coeff_matrix,
         cluster_rows = FALSE,          # keep our custom order
         cluster_cols = FALSE,
         na_col = "grey90",             # not significant in that cancer
         color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
         main = "Linear regression coefficients (stage ~ embedding feature)\nOrdered by mean coefficient (FDR < 0.2 in at least one cancer)",
         fontsize_row = 6,
         angle_col = 45,
         legend_title = "Coefficient β")

```

# Survival Cox univariate

```{r}
kirc_meta <- lung$meta
names(kirc_meta)
kirc_meta %>%
  select(patient.days_to_last_followup, days_to_last_followup, days_to_death)
```






```{r}
get_survival_coefficients <- function(embed_mat, meta_sub, time_col = "days_to_death", 
                                      event_col = "vital_status", fdr_thresh = 0.2) {
  # embed_mat: patients x features
  # meta_sub: data frame with patient metadata (must contain time and event columns)
  # time_col: column name for survival time (days)
  # event_col: column name for event (1 = death, 0 = censored)
  
  library(survival)
  
  # Align patients
  common <- intersect(rownames(embed_mat), rownames(meta_sub))
  embed <- embed_mat[common, ]
  meta <- meta_sub[common, ]
  
  # Extract time and event
  time <- meta[[time_col]]
  event <- meta[[event_col]]
  
  # Remove patients with missing time or event
  valid <- complete.cases(time, event)
  embed <- embed[valid, ]
  time <- time[valid]
  event <- event[valid]
  
  # Build survival object
  surv_obj <- Surv(time, event)
  
  # Loop over each embedding feature
  coeff <- numeric(ncol(embed))
  pval <- numeric(ncol(embed))
  
  for (i in 1:ncol(embed)) {
    fit <- coxph(surv_obj ~ embed[, i])
    coeff[i] <- coef(fit)[1]           # log hazard ratio
    pval[i] <- summary(fit)$coefficients[1, 5]  # p-value for the coefficient
  }
  
  names(coeff) <- colnames(embed)
  names(pval) <- colnames(embed)
  
  fdr <- p.adjust(pval, method = "fdr")
  sig <- which(fdr < fdr_thresh)
  
  data.frame(
    Feature = names(coeff)[sig],
    LogHR = coeff[sig],
    P_value = pval[sig],
    FDR = fdr[sig],
    stringsAsFactors = FALSE
  )
}
```


```{r}
t = as.data.frame(kirc$meta)
table(lung$meta$vital_status)
# For KIRC
kirc_surv <- get_survival_coefficients(kirc$embed, kirc$meta,
                                       time_col = "days_to_death",
                                       event_col = "vital_status",
                                       fdr_thresh = 0.5)

# For COAD
coad_surv <- get_survival_coefficients(coad$embed, coad$meta,
                                       time_col = "days_to_death.x",
                                       event_col = "vital_status.x",
                                       fdr_thresh = 0.2)

# For LUAD
lung_surv <- get_survival_coefficients(lung$embed, lung$meta,
                                       time_col = "days_to_death",
                                       event_col = "vital_status",
                                       fdr_thresh = 0.2)

# Add cancer labels
kirc_surv$Cancer <- "KIRC"
coad_surv$Cancer <- "COAD"
lung_surv$Cancer <- "LUAD"


```


```{r}
n_top <- 30

# KIRC
kirc_top_s <- kirc_surv[order(-abs(kirc_surv$LogHR)), ][1:n_top, ]
kirc_top_s <- kirc_top_s[order(-kirc_top_s$LogHR), ]
mat_kirc_s <- matrix(kirc_top_s$LogHR, ncol = 1,
                     dimnames = list(kirc_top_s$Feature, "KIRC"))

# COAD
coad_top_s <- coad_surv[order(-abs(coad_surv$LogHR)), ][1:n_top, ]
coad_top_s <- coad_top_s[order(-coad_top_s$LogHR), ]
mat_coad_s <- matrix(coad_top_s$LogHR, ncol = 1,
                     dimnames = list(coad_top_s$Feature, "COAD"))

# LUAD
lung_top_s <- lung_surv[order(-abs(lung_surv$LogHR)), ][1:n_top, ]
lung_top_s <- lung_top_s[order(-lung_top_s$LogHR), ]
mat_lung_s <- matrix(lung_top_s$LogHR, ncol = 1,
                     dimnames = list(lung_top_s$Feature, "LUAD"))

# Create heatmaps
p_kirc_s <- pheatmap(mat_kirc_s, cluster_rows = FALSE, cluster_cols = FALSE,
                     color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                     fontsize_row = 8, legend_title = "Log HR", silent = TRUE)
p_coad_s <- pheatmap(mat_coad_s, cluster_rows = FALSE, cluster_cols = FALSE,
                     color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                     fontsize_row = 8, legend_title = "Log HR", silent = TRUE)
p_lung_s <- pheatmap(mat_lung_s, cluster_rows = FALSE, cluster_cols = FALSE,
                     color = colorRampPalette(c("steelblue", "white", "firebrick3"))(50),
                     fontsize_row = 8, legend_title = "Log HR", silent = TRUE)

grid.arrange(p_kirc_s[[4]], p_coad_s[[4]], p_lung_s[[4]], ncol = 3,
             top = "Top embedding features associated with survival (FDR < 0.2)")
```




# Compute pathway scores (GSVA) and correlate with embeddings


```{r}
get_gsva_pathway <- function(mae_emb, rna_assay_name, aligned_list, res, top_n = NULL) {
  # mae_emb: MultiAssayExperiment from linkTCGA
  # rna_assay_name: character, e.g. "KIRC_RNASeq2GeneNorm-20160128"
  # aligned_list: list with $embed (patients x features) and $meta
  # res: data frame from get_stage_coefficients (with Feature, Coefficient, FDR)
  # top_n: optional, number of top stage‑associated features to use (by |Coefficient|)
  
  # 1. Get Hallmark gene sets (cached)
  hallmark <- msigdbr(species = "Homo sapiens", collection = "H") %>%
    split(x = .$gene_symbol, f = .$gs_name)
  
  # 2. Extract RNA and align to patients (patients from aligned_list$embed)
  rna_se <- mae_emb[[rna_assay_name]]
  rna_mat <- assay(rna_se)  # genes x samples
  
  sm <- sampleMap(mae_emb)
  sm_rna <- sm[sm$assay == rna_assay_name, ]
  patients <- rownames(aligned_list$embed)
  
  rna_patient <- matrix(NA, nrow = nrow(rna_mat), ncol = length(patients),
                        dimnames = list(rownames(rna_mat), patients))
  for (pat in patients) {
    cols <- sm_rna$colname[sm_rna$primary == pat]
    if (length(cols) > 0) {
      rna_patient[, pat] <- rowMeans(rna_mat[, cols, drop = FALSE], na.rm = TRUE)
    }
  }
  keep <- rowSums(!is.na(rna_patient)) > 0
  rna_patient <- rna_patient[keep, , drop = FALSE]
  
  # 3. GSVA
  ssgsea_param <- ssgseaParam(exprData = rna_patient, geneSets = hallmark,
                              alpha = 0.25, normalize = TRUE)
  gsva_scores <- gsva(ssgsea_param, verbose = FALSE)   # pathways x patients
  pathway_scores <- t(gsva_scores)                     # patients x pathways
  
  # 4. Align with embeddings
  common <- intersect(rownames(aligned_list$embed), rownames(pathway_scores))
  embed_aligned <- aligned_list$embed[common, ]
  pathway_aligned <- pathway_scores[common, ]
  
  # 5. Correlation (features x pathways)
  cor_embed_pathway <- cor(embed_aligned, pathway_aligned, use = "pairwise.complete.obs")
  
  # 6. Select stage‑associated features (from regression results)
  if (!is.null(top_n) && top_n < nrow(res)) {
    stage_features <- head(res[order(-abs(res$Coefficient)), "Feature"], top_n)
  } else {
    stage_features <- res$Feature
  }
  
  # 7. Subset correlation matrix
  pathway_cor_subset <- cor_embed_pathway[stage_features, , drop = FALSE]
  
  # 8. For each feature, get top 3 positive and negative pathways
  pathway_summary <- list()
  for (feat in stage_features) {
    cors <- pathway_cor_subset[feat, ]
    pos <- sort(cors, decreasing = TRUE)[1:3]
    neg <- sort(cors, decreasing = FALSE)[1:3]
    pathway_summary[[feat]] <- data.frame(
      Embedding_Feature = feat,
      Direction = rep(c("Positive", "Negative"), each = 3),
      Pathway = c(names(pos), names(neg)),
      Correlation = c(pos, neg),
      stringsAsFactors = FALSE
    )
  }
  pathway_df <- do.call(rbind, pathway_summary)
  rownames(pathway_df) <- NULL
  
  return(pathway_df)
}
```


```{r}
# KIRC
pathway_kirc <- get_gsva_pathway(
  mae_emb = kidney_mae_emb,
  rna_assay_name = "KIRC_RNASeq2GeneNorm-20160128",
  aligned_list = kirc,
  res = kirc_res,
  top_n = 30   # use top 30 stage features (by |coefficient|)
)

# COAD
pathway_coad <- get_gsva_pathway(
  mae_emb = coad_mae_emb,
  rna_assay_name = "COAD_RNASeq2GeneNorm-20160128",
  aligned_list = coad,
  res = coad_res,
  top_n = 30
)

# LUAD
pathway_luad <- get_gsva_pathway(
  mae_emb = lung_mae_emb,
  rna_assay_name = "LUAD_RNASeq2Gene-20160128",
  aligned_list = lung,
  res = lung_res,
  top_n = 30
)
```


